Metric Spaces with Expensive Distances

### 2019-04-17 paper/expensive-distances-main.tex
add timings to paper

\section{Experiments on spanners}
We run experiments on the points sampled from the low-dimensional Euclidean space to investigate
experimentally the performance of these heuristics. Clearly, for this metric space, our
cost model is not meaningful since distance comparisons are cheap;
but we picked this environment for controlled experiments.
In order to test the \textsc{BlindQuasiSorted} algorithms we multiply the true distance by a factor from $[1,2]$ chosen uniformly.

@@

\end{axis}
\end{tikzpicture}
\end{centering}
    \caption{Number of edges in blind spanners generated by different variants of the blind
    algorithm. Greedy non-blind algorithm and WSPD algorithm are included for comparison. The plot is for normally distributed points
    in dimension 2, $\eps = 0.1$.}
    \label{fig:spanner_sparseness}
\end{figure}

\begin{figure}[ht]
    \begin{centering}
\begin{tikzpicture}
    \begin{axis}[xlabel=\# Points,ylabel=\# Edges / \# Points,
        legend pos=outer north east]

@@

\end{axis}
\end{tikzpicture}
\end{centering}
    \caption{Ratio \# edges / \# points for different variants of spanner 
    algorithms. The plot is for normally distributed points
    in dimension 2, $\eps = 0.1$.}
    \label{fig:spanner_ratio}
\end{figure}

\begin{figure}[ht!]
    \includegraphics[width=0.7\textwidth]{\expDistFigPath blind-greedy-dims.png}
    \caption{Results of blind greedy spanner for different dimensions.}
    \label{fig:blind_greedy_dimensions}
\end{figure}

\ignore{
\begin{figure}[ht!]
    \begin{centering}
\begin{tikzpicture}
    \begin{axis}[xlabel=\# Points,ylabel=\# Edges,
        legend pos=outer north east]

\end{axis}
\end{tikzpicture}
\end{centering}
    \caption{Comparison of the four variants of \brndm algorithm.}
    \label{fig:blind_rbr_variants}
\end{figure}
}

\begin{figure}[ht!]
    \begin{centering}
\begin{tikzpicture}
    \begin{axis}[xlabel=epsilon,ylabel=\# Edges,
        legend pos=outer north east]

\end{axis}
\end{tikzpicture}
\end{centering}
    \caption{Number of edges in the blind greedy and greedy spanners for different values of $\eps$. Data is for 400 normally distributed points in $\RR^2$ and $\RR^3$.}
    \label{fig:spanner_eps_dependence}
\end{figure}

@@

\end{axis}
\end{tikzpicture}
\end{centering}
    \caption{Sparseness of \bgrdy spanner algorithm. The plot is for 0-dimensional diagrams of perturbed McGill data
    $\eps = 0.1, 0.2, 0.5$ with $W_1$ distance.}
    \label{fig:spanner_sparseness_mcgill_perturbed}
\end{figure}

\begin{table}\centering
    \ra{1.2}
    \begin{tabular}{@{}rrrcrrcrr@{}}\toprule
                          & \multicolumn{2}{c}{$q=1$} & \phantom{abc} & \multicolumn{2}{c}{$q=2$} & \phantom{abc} & \multicolumn{2}{c}{$q=3$}\\
                          \cmidrule{2-3} \cmidrule{5-6} \cmidrule{8-9}
                          & Greedy & Blind greedy && Greedy & Blind greedy && Greedy & Blind greedy \\ \midrule
$\eps = 0.1$\\
%\#edges                    & 12755 & 28350         && 32606 & 72933      &&  41875  & 78977 \\
\% of distances computed   & 13.63 & 28.06         && 32.27 & 72.19      &&   41.45  & 78.17 \\
$\eps = 0.2$\\
%\#edges                    & 6314 & 19518         && 13560 & 44282       &&  20335  & 52482 \\
\% of distances computed   & 6.25 & 19.32         && 13.42 & 43.83       &&   20.13  & 51.95 \\
$\eps = 0.5$\\
%\#edges                    & 2263 & 10414         && 3411 & 20143        &&  4705  & 27154 \\
\% of distances computed   & 2.24 & 10.3          && 3.38 & 19.94        &&   4.66  & 26.88 \\
\bottomrule
\end{tabular}
\caption{Comparison of \bgrdy and \textsc{Greedy} algorithms for 0-dimensional
diagrams of shapes from McGill dataset with Wasserstein distance $W_q$. The number of diagrams is 450.}
\label{tbl:mcgill_original_blind_greedy_spanner}
\end{table}

We tested the algorithm for $\eps \in \Set{0.01, 0.1, 0.2, 0.5}$ on the following sets of points in dimensions $d = 2,3,4,5$:
\begin{enumerate}
    \item In the \textbf{uniform} test set points are sampled uniformly at
        random from the unit cube in $\mathbb{R}^d$.
    \item In the \textbf{normal} test set points are sampled from the standard
        normal distribution in $\mathbb{R}^d$.
    \item In the \textbf{clustered} test set we first sample cluster centers uniformly 
        at random from $[0,10000]^d$, and then we add normally distributed noise around
        each of the centers. The number of clusters is chosen so that each cluster
        contains 50 points.
    \item The test set \textbf{exp} consists of points of the form $(2^{\xi_1}, \dots, 2^{xi_d})$,
        where $\xi_i$'s are i.i.d. random variables with uniform distribution on $[1,25]$.
\end{enumerate}
In all experiments the algorithms that we tested compared in the same way,
so we only present results for the \textbf{uniform} point set in dimension 2.

\cref{fig:spanner_sparseness}
shows the number of edges of the spanner for various variants of 
blind and non-blind spanner constructions.
Note that for all blind spanner variants, the number of computed distances
is equal to the spanner size, while for the non-blind greedy spanner,
this number is always $\binom{n}{2}$ and for WPSD it is lower bounded
by the size of the spanner.
We can see that, even though none of the blind spanners can produce
spanners of the same quality (i.e., sparse) as the standard greedy algorithm,
\bgrdy and all variants of \brndm
perform significantly better than both variants of \textsc{BlindQuasiSorted}.
\cref{fig:spanner_ratio} shows the ratio of the number of edges to the number of points.
The ideal behavior is demonstrated by the non-blind greedy spanner,
for which this ratio stays practically constant, confirming the linear growth.
None of the blind algorithms seems to have this property, but among them
the blind greedy spanner is the best one. If we assume that the number of edges
is proportional to $n^\alpha$, then we can try to estimate $\alpha$ by 
linear regression (after taking $\log$).  We give in \cref{tbl:regr_coeff_spanner}
the estimated exponents $\alpha$ for \bgrdy and standard
greedy algorithms. Note that even for the greedy algorithm these estimated
exponents can be significantly larger than 1,
which is explained by the fact that the number of points
on which we computed spanners is not large enough to clearly see
the linear dependence.

\begin{table}[]
\begin{tabular}{|l|l|l|}
\hline
dimension & \textbf{Greedy (non-blind)} & \textbf{Blind greedy} \\ \hline
2         &         1.08                &  1.12                 \\ \hline
3         &         1.24                &  1.41                \\ \hline
4         &         1.42                &  1.77                 \\ \hline
\end{tabular}
\caption{Estimated exponents in the $|E|= C |V|^\alpha$ dependence of the number of edges
on the number of points. The data is for $\eps = 0.1$ and for uniform points.}
\label{tbl:regr_coeff_spanner}
\end{table}

@@

\ignore{
Summing up, the \bgrdy algorithm is the best algorithm for our model, but also \brndm
algorithm reduces the amount of computed distances substantially, especially if we enforce having non-zero lower bounds
first. 
If the goal is to reduce the number of distance computations, these method seem to be more suitable than a WSPD spanner.
Since the linear spanner size of WSPDs does not show up in the experiments because of the relatively small values of $n$
tested, the experiments are not conclusive regarding the asymptotic size of the blind spanners.
Another noteworthy fact is that quasi-sorted variants produce spanners which are much closer to the complete graph
(\bqsgrdy is worse, requiring all the edges). It would seem plausible that, if we have
access to approximate value of the distance, we could exploit this in the spanner construction, but we could not
find a working heuristic.
}

From these experiments we conclude that the algorithm with the highest chance
of saving a significant amount of distance computations on real data, like large
persistence diagrams, is \bgrdy. Even though \brdnm also substantially reduces
the number of computed distances, its performance is significantly worse;
other algorithms that we tested do not produce sparse spanners.
While it would seem plausible that access to approximate value of the distance 
could be exploited in the spanner construction, 
we could not find a working heuristic; \bqsgrdy and \bqsshaker produce extremely
dense spanners.

@@

Results of \bgrdy algorithm on the whole \dtype{original} dataset 
for different values of $\eps$ and $q$ are summarized in \cref{tbl:mcgill_original_blind_greedy_spanner};
\grdy algorithm is used as baseline\footnote{Of course, in the columns of \grdy algorithm
we give its sparseness, i.e. number of edges divided by the number of all distances;
the percent of computed distances is always 100.}.  The first thing to note is that for higher values of $q$
any advantage of our algorithm disappears. However, the cardinality of the \grdy
spanner is also increasing; for instance, if we want to achieve 10 \% accuracy,
then even the \grdy spanner is almost half of the complete graph.
We can offer the following explanation, assuming the reader is familiar
with the definition of $\wsdist{q}$. For higher values of $q$,
longer edges of an optimal matching play a more important role
than shorter ones; as $q$ tends to $\infty$, Wasserstein distance
approaches bottleneck distance $\btdist$, which depends on the longest
edge only. If all diagrams come from the same dataset,
then there is a large number of pairs of diagrams whose optimal matchings
have the longest edges of approximately the same length. This fact means that we are in the case
of high doubling dimension, being close to a clique, and there is no hope
of getting a small spanner. \ignore{On the other hand, computation of $\btdist$ in practice
requires significantly less time than computation of $\wsdist{q}$, and the case
of $\wsdist{1}$ and $\wsdist{2}$ is of practical importance.}
For $\wsdist{1}$ we observe more encouraging figures, like avoiding 70 \% of distance
computations for $\eps = 0.1$; if we are happy with a 1.5 approximation, than we
reduce the number of distances computed by a factor of 10.

@@

The proof is given in the appendix.

@@

\appendix
\section{Proof of \cref{thm:ann_bound}}

>>>>

\section{Experiments on spanners}
We run experiments on the points sampled from the low-dimensional Euclidean space to investigate
experimentally the performance of these heuristics. Clearly, for this metric space, our
cost model is not meaningful since distance comparisons are cheap;
but we picked this environment for controlled experiments.
In order to test the \textsc{BlindQuasiSorted} algorithms we multiply the true distance by a factor from $[1,2]$ chosen uniformly.
By \textit{sparseness} of a graph with $n$
vertices we always mean the number of edges of the graph divided by $\binom{n}{2}$, the number of edges in the complete graph on $n$ edges.

@@

\end{axis}
\end{tikzpicture}
\end{centering}
    \caption{Number of edges in blind spanners generated by different variants of the blind
    algorithm. Greedy non-blind algorithm and WSPD algorithm are included for comparison. The plot is for uniformly distributed points
    in dimension 2, $\eps = 0.1$.}
    \label{fig:spanner_sparseness}
\end{figure}

\begin{figure}[ht]
        \begin{tikzpicture}
            \begin{axis}[xlabel=\# Points,ylabel=\# Edges / \# Points,
                legend pos=outer north east]

@@

            \end{axis}
        \end{tikzpicture}
            \caption{Ratio \# edges / \# points for different variants of spanner 
            algorithms. The plot is for normally distributed points
            in dimension 2, $\eps = 0.1$.}
            \label{fig:spanner_ratio}
\end{figure}

@@

\end{axis}
\end{tikzpicture}
\end{centering}
    \caption{Sparseness of \bgrdy spanner algorithm. The plot is for \dtype{perturbed} persistence diagrams,
    $\eps = 0.1, 0.2, 0.5$ with $W_1$ distance.}
    \label{fig:spanner_sparseness_mcgill_perturbed}
\end{figure}

\begin{table}\centering
    \ra{1.2}
    \begin{tabular}{@{}rrrcrrcrr@{}}\toprule
                          & \multicolumn{2}{c}{$q=1$} & \phantom{abc} & \multicolumn{2}{c}{$q=2$} & \phantom{abc} & \multicolumn{2}{c}{$q=3$}\\
                          \cmidrule{2-3} \cmidrule{5-6} \cmidrule{8-9}
                          & Greedy & Blind greedy && Greedy & Blind greedy && Greedy & Blind greedy \\ \midrule
%\#edges                    & 12755 & 28350         && 32606 & 72933      &&  41875  & 78977 \\
$\eps = 0.1$                & 0.14  & 0.28         && 0.32  & 0.72      &&   0.41  & 0.78 \\
%\#edges                    & 6314 & 19518         && 13560 & 44282       &&  20335  & 52482 \\
$\eps = 0.2$                & 0.06 & 0.19          && 0.13 & 0.44       &&   0.20  & 0.52 \\
%\#edges                    & 2263 & 10414         && 3411 & 20143        &&  4705  & 27154 \\
%\% of distances computed   & 2.24 & 10.3          && 3.38 & 19.94        &&   4.66  & 26.88 \\
$\eps = 0.5$               & 0.02 &  0.10          && 0.03 & 0.20        &&   0.05  & 0.27 \\
\bottomrule
\end{tabular}
    \caption{Sparseness of \bgrdy and \textsc{Greedy} spanners for 450 \dtype{original} persistence diagrams
 with Wasserstein distance $W_q$.}
\label{tbl:mcgill_original_blind_greedy_spanner_sparseness}
\end{table}

\begin{table}\centering
    \ra{1.2}
    \begin{tabular}{@{}rrrcrrcrr@{}}\toprule
                          & \multicolumn{2}{c}{$q=1$} & \phantom{abc} & \multicolumn{2}{c}{$q=2$} & \phantom{abc} & \multicolumn{2}{c}{$q=3$}\\
                          \cmidrule{2-3} \cmidrule{5-6} \cmidrule{8-9}
                          & Bruteforce & Blind greedy && Bruteforce & Blind greedy && Bruteforce & Blind greedy \\ \midrule

$\eps = 0.1$    &   4093  &  1266  &&  5331  &  4085  &&  16852  &  13557\\
$\eps = 0.2$    &   4093  &  878  &&  5331  &  2574  &&  16852  &  9383\\
$\eps = 0.5$    &   4093  &  483  &&  5331  &  1233  &&  16852  &  5146\\
\bottomrule
\end{tabular}
    \caption{Time required to approximate all pairwise distances with \bgrdy algorithm.
    Data is for 450 \dtype{original} persistence diagrams with Wasserstein distance $W_q$, time in seconds.}
\label{tbl:mcgill_original_blind_greedy_spanner_timing}
\end{table}

We tested the algorithms for $\eps \in \Set{0.01, 0.1, 0.2, 0.5}$ on the following sets of points in dimensions $d = 2,3,4,5$:
\begin{enumerate}
    \item In the \textbf{uniform} test set points are sampled uniformly at
        random from the unit cube in $\mathbb{R}^d$.
    \item In the \textbf{normal} test set points are sampled from the standard
        normal distribution in $\mathbb{R}^d$.
    \item In the \textbf{clustered} test set we first sample cluster centers uniformly 
        at random from $[0,10000]^d$, and then we add normally distributed noise around
        each of the centers. The number of clusters is chosen so that each cluster
        contains 50 points.
    \item The test set \textbf{exp} consists of points of the form $(2^{\xi_1}, \dots, 2^{xi_d})$,
        where $\xi_i$'s are i.i.d. random variables with uniform distribution on $[1,25]$.
\end{enumerate}
In all experiments the algorithms that we tested compared in the same way,
so we only present results for the \textbf{uniform} point set in dimension 2.

\cref{fig:spanner_sparseness}
shows the number of edges of the spanner for various variants of 
blind and non-blind spanner constructions.
Note that for all blind spanner variants, the number of computed distances
is equal to the spanner size, while for the non-blind greedy spanner,
this number is always $\binom{n}{2}$ and for WPSD it is lower bounded
by the size of the spanner.
We can see that, even though none of the blind spanners produces
spanners of the same quality as the standard greedy algorithm,
\bgrdy and all variants of \brndm
perform significantly better than both variants of \textsc{BlindQuasiSorted}.

\ignore{
As for different variants of the \brndm algorithm,
we note that their performance is almost the same, and the algorithm works
significantly better than \textsc{QuasiSorted} variants, but obviously
worse than the blind greedy variant. There is a consistent, though small,
difference between the variants that do not force lower bounds first and
the other two variants of the \brndm.  (see \cref{fig:blind_rbr_variants}). 
}

@@

%\ignore{
%Summing up, the \bgrdy algorithm is the best algorithm for our model, but also \brndm
%algorithm reduces the amount of computed distances substantially, especially if we enforce having non-zero lower bounds
%first. 
%If the goal is to reduce the number of distance computations, these method seem to be more suitable than a WSPD spanner.
%Since the linear spanner size of WSPDs does not show up in the experiments because of the relatively small values of $n$
%tested, the experiments are not conclusive regarding the asymptotic size of the blind spanners.
%Another noteworthy fact is that quasi-sorted variants produce spanners which are much closer to the complete graph
%(\bqsgrdy is worse, requiring all the edges). It would seem plausible that, if we have
%access to approximate value of the distance, we could exploit this in the spanner construction, but we could not
%find a working heuristic.
%}

From these experiments we conclude that the algorithm with the highest chance
of saving a significant amount of distance computations on real data, like large
persistence diagrams, is \bgrdy. Even though \brndm also substantially reduces
the number of computed distances, its performance is significantly worse;
other algorithms that we tested do not produce sparse spanners.
While it would seem plausible that access to approximate value of the distance 
could be exploited in the spanner construction, 
we could not find a working heuristic; \bqsgrdy and \bqsshaker produce extremely
dense spanners.

@@

Results of \bgrdy algorithm on the whole \dtype{original} dataset 
for different values of $\eps$ and $q$ are summarized in \cref{tbl:mcgill_original_blind_greedy_spanner_sparseness}
and in \cref{tbl:mcgill_original_blind_greedy_spanner_timing}.
\grdy algorithm is used as baseline.  The first thing to note is that for higher values of $q$
any advantage of our algorithm disappears. However, the cardinality of the \grdy
spanner is also increasing; for instance, if we want to achieve 10 \% accuracy,
then even the \grdy spanner is almost half of the complete graph.
We can offer the following explanation, assuming the reader is familiar
with the definition of $\wsdist{q}$. For higher values of $q$,
longer edges of an optimal matching play a more important role
than shorter ones; as $q$ tends to $\infty$, Wasserstein distance
approaches bottleneck distance $\btdist$, which depends on the longest
edge only. If all diagrams come from the same dataset,
then there is a large number of pairs of diagrams whose optimal matchings
have the longest edges of approximately the same length. This fact means that we are in the case
of high doubling dimension, being close to a clique, and there is no hope
of getting a small spanner. \ignore{On the other hand, computation of $\btdist$ in practice
requires significantly less time than computation of $\wsdist{q}$, and the case
of $\wsdist{1}$ and $\wsdist{2}$ is of practical importance.}
For $\wsdist{1}$ we observe more encouraging figures, like avoiding 70 \% of distance
computations for $\eps = 0.1$; if we are happy with a 1.5 approximation, then we
reduce the number of distances computed by a factor of 10.

@@

The proof is given in the appendix \cref{sec:proof}.

@@

\appendix

\section{Additional Experimental Results on Spanners}
\label{sec:spanners_euclidean_addendum}

\begin{figure}[ht!]
    \begin{centering}
\begin{tikzpicture}
    \begin{axis}[xlabel=\# Points,ylabel=\# Edges,
        legend pos=outer north east]

\end{axis}
\end{tikzpicture}
\end{centering}
    \caption{Comparison of the four variants of \brndm algorithm.}
    \label{fig:blind_rbr_variants}
\end{figure}

\begin{figure}[ht!]
    \begin{centering}
    \includegraphics[width=0.8\textwidth]{\expDistFigPath blind-greedy-dims.png}
    \caption{Results of \bgrdy spanner for different dimensions.}
    \label{fig:blind_greedy_dimensions}
    \end{centering}
\end{figure}

\cref{fig:spanner_ratio} shows the ratio of the number of edges to the number of points.
The ideal behavior is demonstrated by the non-blind greedy spanner,
for which this ratio stays practically constant, confirming the linear growth.
None of the blind algorithms seems to have this property, but among them
the blind greedy spanner is the best one. If we assume that the number of edges
is proportional to $n^\alpha$, then we can try to estimate $\alpha$ by 
linear regression (after taking $\log$).  We give in \cref{tbl:regr_coeff_spanner}
the estimated exponents $\alpha$ for \bgrdy and standard
greedy algorithms. Note that even for the greedy algorithm these estimated
exponents can be significantly larger than 1,
which is explained by the fact that the number of points
on which we computed spanners is not large enough to clearly see
the linear dependence.

\begin{table}[b]
\begin{tabular}{|l|l|l|}
\hline
dimension & \textbf{Greedy (non-blind)} & \textbf{Blind greedy} \\ \hline
2         &         1.08                &  1.12                 \\ \hline
3         &         1.24                &  1.41                \\ \hline
4         &         1.42                &  1.77                 \\ \hline
\end{tabular}
\caption{Estimated exponents in the $|E|= C |V|^\alpha$ dependence of the number of edges
on the number of points. The data is for $\eps = 0.1$ and for uniform points.}
\label{tbl:regr_coeff_spanner}
\end{table}

\begin{figure}[ht]
        \begin{tikzpicture}
            \begin{axis}[xlabel=epsilon,ylabel=\# Edges,
                legend pos=north east]

        \end{axis}
        \end{tikzpicture}
        \caption{Number of edges in the blind greedy and greedy spanners for different values of $\eps$. Data is for 400 normally distributed points in $\RR^2$ and $\RR^3$.}
        \label{fig:spanner_eps_dependence}
\end{figure}

\section{Proof of \cref{thm:ann_bound}}
\label{sec:proof_ann}
